Binocular combination in the autonomic nervous system

### 2025-02-24 manuscript.Rmd
Modified reference style

---
title: 'Binocular combination in the autonomic nervous system'
author: |
 | Federico G. Segala$^1$, Aurelio Bruno$^{1,2}$, Joel T. Martin$^{1,3}$,
 | Anisa Y. Morsi$^1$, Alex R. Wade$^{1,4}$ & Daniel H. Baker$^{1,4}$
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: true
  html_document: default
bibliography: references.bib
link-citations: yes
linkcolor: black
csl: the-journal-of-neuroscience.csl
header-includes:
  \usepackage{caption}
  \usepackage{setspace}\doublespacing
  \usepackage{lineno} \linenumbers
  \newcommand{\beginsupplement}{\setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

$^1$Department of Psychology, University of York, Heslington, York, YO10 5DD, United Kingdom.

$^2$School of Psychology and Vision Sciences, University of Leicester, Leicester, LE1 7RH, United Kingdom.

$^3$School of Philosophy, Psychology and Language Sciences, University of Edinburgh, Edinburgh, EH8 9AD, United Kingdom.

$^4$York Biomedical Research Institute, University of York, Heslington, York, YO10 5NG, United Kingdom.

Corresponding author: Federico G. Segala, [federico.segala@york.ac.uk](federico.segala@york.ac.uk).

@@

The autonomic nervous system regulates many involuntary bodily processes, including the constriction and dilation of the pupils in response to light [@McDougal2015]. The anatomical pathway from the retina to the subcortical nuclei controlling the pupillary light response (PLR) is well established: it includes the Pretectal Olivary nucleus (PON), the Superior Cervical ganglion and the Edinger-Westphal nucleus, which project to the iris sphincter muscles that directly control the pupil size [@McDougal2015]. Evidence of a binocular component to the PLR is shown by the consensual response of the pupil [stimulation of one eye will cause constriction of the other eye\; @Wyatt1981]. The anatomical segregation of the subcortical pathway from the rest of the brain means that this binocular combination of signals must occur independently of the cortical processes of binocular integration required for visual perception. Our recent work [@Segala2023] has shown that the algorithm underlying binocular combination of light in the pupil pathway differs from that in the cortex. Here we extend this paradigm to compare binocular combination in different photoreceptor pathways that feed into the autonomic nervous system.

Different classes of retinal photoreceptors, including cones, rods and melanopsin-containing intrinsically photosensitive retinal ganglion cells (ipRGCs), are directly involved in controlling and maintaining the size of the pupils [e.g. @McDougal2010; @Barrionuevo2016; @Murray2018; @Spitschan2014; @Woelders2018]. Cones drive the initial rapid constriction of the pupils [@Mathot2018], while the slower and longer activation of the ipRGCs maintains constriction over a prolonged period of time and regulates the post-illumination pupillary response [@Markwell2010; @McDougal2010]. The ipRGCs are a recently discovered photoreceptor class [@Provencio2000] that express the photopigment melanopsin, and are involved in the regulation of the circadian rhythm [@Panda2002; @Ruby2002], forming a major input to the PON [@Dacey2003]. The first direct evidence of the involvement of the ipRGCs in the PLR was shown in melanopsin knockout rats by @Lucas2003, resulting in the loss of the intrinsic photosensitivity of the cells and a reduced pupil constriction. Similar behaviour was later observed in primates and humans [@Gamlin2007] using silent substitution (see Methods), where it was demonstrated that the PLR continues during light presentation even when cone and rod signalling is blocked, indicating the primary role of the ipRGCs in maintaining pupil constriction over a prolonged time.

Binocular combination has been extensively studied in visual perception, where it is mediated by neurons in primary visual cortex [@Hubel1962]. For pattern vision, binocular summation occurs at threshold, such that a lower contrast is required to detect a target shown to both eyes than a target shown to one eye [@Baker2018; @Campbell1965]. At higher contrasts, the phenomenon of 'ocularity invariance' is observed, in which the response to monocularly- and binocularly-presented patterns is equal [@Baker2007; @Moradi2009]. This is explained by a process of interocular suppression that cancels out the additional excitatory drive caused by stimulating two eyes. Our recent work [@Segala2023] showed that cortical signal combination for luminance flicker is substantially more linear than for spatial patterns, whereas there is evidence of interocular suppression in the pupil pathway. Here we measure the amplitude of pupil modulations in response to flickering stimuli presented as light flux, or directed towards the L-M cone, S-cone, and melanopsin pathways (see Figure \@ref(fig:spectraplots)e-l). Our key comparisons are between monocular and binocular stimulation, and in a dichoptic masking condition where the two eyes are stimulated at different frequencies. The results are interpreted using a contemporary model of binocular vision [@Meese2006] implemented within a hierarchical Bayesian framework.

@@

Twenty-four participants were recruited for each of the four experiments for a total of ninety-six adult participants (28 male, 68 female), whose ages ranged from 18 to 41. All participants had normal or corrected to normal vision, no known abnormalities of binocular or colour vision, and gave written informed consent. Our procedures were approved by the Ethics Committee of the Department of Psychology at the University of York (identification number 184).

@@

When preparing the spectral sequences, the age of participants was used to account for the yellowing of the lenses. We used the silent substitution technique [@Estevez1982] to selectively stimulate specific photoreceptor classes. Silent substitution exploits the fact that each photoreceptor class has a distinct spectral tuning that overlaps with the others. Using a multiprimary system, in which the primaries (i.e. LEDs) have different spectra, it is possible to target one class of photoreceptors while maintaining the others at a constant activity level, effectively silencing them [@Shapiro1996; @Spitschan2018, this paper also offers a clear explanation of how to implement silent substitution]. We calculated silent substitution solutions using the _PySilSub_ toolbox [@Martin2023], using linear algebra. The outputs of the two light engines (see Figure \@ref(fig:spectraplots)b,c) were calibrated using an Ocean Optics Jaz spectroradiometer, which was wavelength-calibrated to an Argon lamp and intensity calibrated using a NIST-traceable light source. Each primary was also linearized using a polynomial fit (see Figure \@ref(fig:primarylinearity) for details). We used the @Stockman2000 10-degree cone fundamentals, and estimates of melanopsin absorbance spectra from CIE S 026 [discussed in @Martin2023] to calculate $\alpha$-opic irradiance.

@@

The diffuser discs were positioned in the objective planes of the lenses of a modified VR headset (SHINECON SC-G01, Dongguan Shinecon Industrial Co. Ltd., Guangdong, China), which was used by the participants to view the stimuli. The stimuli were two discs of flickering light with a diameter of approximately 30 degrees, which were fused together into a cyclopean percept resembling a donut-shaped ring of light, similar to that used in other studies [e.g., @Barrionuevo2016; @Murray2018; @Spitschan2014; @Spitschan2015; @Zele2018]. The VR headset modifications allowed for small adjustments to account for individual differences in interpupillary distance and focal length. The use of this set up allowed us to modulate the stimuli in three different ocular configurations, similar to the ones we used in our previous study [@Segala2023]: monocular, binocular and dichoptic. In the monocular configuration, the unstimulated eye still saw a non-flickering disc of mean light flux. A schematic of the stimulation system is shown in Figure \@ref(fig:spectraplots)a. Pupillometry data were collected using a binocular Pupil Core eye-tracker headset [Pupil Labs GmbH, Berlin, Germany, @Kassner2014] running at 120 Hz, and the signals were recorded with the Pupil Capture software.

Our previous study [@Segala2023] used a temporal frequency of 2Hz for foveal luminance flicker, and recorded EEG data simultaneously with pupillometry. Initial pilot experiments indicated that this frequency was too high to elicit measurable responses when stimulating individual photoreceptor pathways. For all experiments, we therefore used a primary flicker frequency of 0.5Hz, as previous literature showed that this was slow enough to elicit a pupil response from all photoreceptor classes [@Spitschan2014]. We also focussed on only recording pupillometry data as this frequency would be too slow to elicit steady-state EEG responses [@Norcia2015].

@@

Before the start of each experiment, participants adjusted the objective planes of the lenses with the help of the experimenter until the stimulus was in focus and they perceived the two pieces of blackout material as one fused disc. Pupil responses to binocular temporal contrast modulations were examined in a factorial design that combined six ocular conditions and five temporal contrast levels: 6, 12, 24, 48 and 96% of the available dynamic range. This design, similar to that used in our previous studies [@Segala2023; @Baker2017; @Baker2020], was applied in four separate experiments, each with a different mode of photoreceptor stimulation. In the first three conditions, the discs flickered at 0.5 Hz, in either a monocular, binocular or dichoptic arrangement. In the dichoptic condition the non-target eye saw a flickering fixed contrast of 48% of the available dynamic range. In the remaining three conditions (the cross-frequency conditions) one eye’s disc flickered at 0.4 Hz, and the other eye’s disc flickered at 0.5 Hz. This included monocular responses at 0.4 Hz, as well as binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 0.5 Hz, mask contrast of 48% at 0.4 Hz in the other eye) arrangements. We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

The pupillometry data were analysed using the same method we used in our previous study [@Segala2023]. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. We used coherent averaging and at each stage we excluded data points with a Mahalanobis distance exceeding $D$ = `r SDthresh` from the complex-valued mean [@Baker2021]. For monocular stimulation, we confirmed that the consensual response was equivalent to the response in the stimulated eye.

@@

To quantitatively summarise our data, we used the same model described in our previous study [@Segala2023]. The model has the same general form as the first stage of the contrast gain control model proposed by @Meese2006, but omits the second stage. For the previous model that we used [@Segala2023], the exponent of the numerator and denominator had fixed values of 2 and (implicitly) 1. Here, we allow these parameters (called _p_ and _q_) to be free, in order to permit different shapes of contrast response function, e.g. accelerating or saturating. The responses of the left eye and right eye channels are as follows:

@@

Our recent psychophysical work [@Baker2024] has looked at binocular interactions in the L-M and S-cone pathways, and compared these to the light flux pathway. Using a contrast discrimination paradigm with spatial modulations of light flux and colour, @Baker2024 found equally strong interocular suppression in all three pathways. This is rather different from our pupillometry results here, which demonstrate weaker suppression in the light flux and S-cone pathways than in the L-M pathway (see Figure \@ref(fig:FirstHarmonicPlots)). However, the current experiments involve temporal modulations, which are quite distinct from the spatial modulations used in @Baker2024. Our other recent work [@Segala2023] has shown that temporal luminance modulations involve much weaker interocular suppression in the cortical response than do spatial luminance modulations [@Baker2017]. These different normalization processes might reflect different priorities for spatial and temporal vision. Spatial vision aims to fuse images to provide binocular single vision, and benefits from 'ocularity invariance' [@Baker2007], in which visual appearance is constant when viewed with one eye or two. Temporal vision is critical for motion perception, which can involve alternative binocular computations, such as calculating velocity differences between the eyes [@Kaestner2019]. Of course the present measurements were of pupil size, which may be subject to different anatomical and functional constraints from those in the cortex. Future research should aim to extend our current findings to perception using psychophysical approaches.

@@

# Acknowledgements

Supported by Biotechnology and Biological Sciences Research Council grant BB/V007580/1 awarded to DHB and ARW.

# Declaration of interests

The authors declare no competing interests.

>>>>

---
title: 'Binocular combination in the autonomic nervous system'
output:
  bookdown::pdf_document2:
    fig_caption: yes
    toc: no
    keep_tex: true
  html_document: default
bibliography: references.bib
link-citations: yes
linkcolor: black
csl: proceedings-of-the-royal-society-b.csl
header-includes:
  \usepackage{caption}
  \usepackage{lineno} \linenumbers
  \newcommand{\beginsupplement}{\setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}}
---

@@

@@

The autonomic nervous system regulates many involuntary bodily processes, including the constriction and dilation of the pupils in response to light [@McDougal2015]. The anatomical pathway from the retina to the subcortical nuclei controlling the pupillary light response (PLR) is well established: it includes the Pretectal Olivary nucleus (PON), the Superior Cervical ganglion and the Edinger-Westphal nucleus, which project to the iris sphincter muscles that directly control the pupil size [@McDougal2015]. Evidence of a binocular component to the PLR is shown by the consensual response of the pupil (stimulation of one eye will cause constriction of the other eye @Wyatt1981). The anatomical segregation of the subcortical pathway from the rest of the brain means that this binocular combination of signals must occur independently of the cortical processes of binocular integration required for visual perception. Recent work [@Segala2023] has shown that the algorithm underlying binocular combination of light in the pupil pathway differs from that in the cortex. Here we extend this paradigm to compare binocular combination in different photoreceptor pathways that feed into the autonomic nervous system.

Different classes of retinal photoreceptors, including cones, rods and melanopsin-containing intrinsically photosensitive retinal ganglion cells (ipRGCs), are directly involved in controlling and maintaining the size of the pupils [e.g. @McDougal2010; @Barrionuevo2016; @Murray2018; @Spitschan2014; @Woelders2018]. Cones drive the initial rapid constriction of the pupils [@Mathot2018], while the slower and longer activation of the ipRGCs maintains constriction over a prolonged period of time and regulates the post-illumination pupillary response [@Markwell2010; @McDougal2010]. The ipRGCs are a recently discovered photoreceptor class [@Provencio2000] that express the photopigment melanopsin, and are involved in the regulation of the circadian rhythm [@Panda2002; @Ruby2002], forming a major input to the PON [@Dacey2003]. The first direct evidence of the involvement of the ipRGCs in the PLR was shown in melanopsin knockout rats [@Lucas2003], resulting in the loss of the intrinsic photosensitivity of the cells and a reduced pupil constriction. Similar behaviour was later observed in primates and humans [@Gamlin2007] using silent substitution (see Methods), where it was demonstrated that the PLR continues during light presentation even when cone and rod signalling is blocked, indicating the primary role of the ipRGCs in maintaining pupil constriction over a prolonged time.

Binocular combination has been extensively studied in visual perception, where it is mediated by neurons in primary visual cortex [@Hubel1962]. For pattern vision, binocular summation occurs at threshold, such that a lower contrast is required to detect a target shown to both eyes than a target shown to one eye [@Baker2018; @Campbell1965]. At higher contrasts, the phenomenon of 'ocularity invariance' is observed, in which the response to monocularly- and binocularly-presented patterns is equal [@Baker2007; @Moradi2009]. This is explained by a process of interocular suppression that cancels out the additional excitatory drive caused by stimulating two eyes. Recent work [@Segala2023] showed that cortical signal combination for luminance flicker is substantially more linear than for spatial patterns, whereas there is evidence of interocular suppression in the pupil pathway. Here we measure the amplitude of pupil modulations in response to flickering stimuli presented as light flux, or directed towards the L-M cone, S-cone, and melanopsin pathways (see Figure \@ref(fig:spectraplots)e-l). Our key comparisons are between monocular and binocular stimulation, and in a dichoptic masking condition where the two eyes are stimulated at different frequencies. The results are interpreted using a contemporary model of binocular vision [@Meese2006] implemented within a hierarchical Bayesian framework.

@@

Twenty-four participants were recruited for each of the four experiments for a total of ninety-six adult participants (28 male, 68 female), whose ages ranged from 18 to 41. All participants had normal or corrected to normal vision, no known abnormalities of binocular or colour vision, and gave written informed consent. Our procedures were approved by the Ethics Committee of the Department of Psychology at the XXX (identification number 184).

@@

When preparing the spectral sequences, the age of participants was used to account for the yellowing of the lenses. We used the silent substitution technique [@Estevez1982] to selectively stimulate specific photoreceptor classes. Silent substitution exploits the fact that each photoreceptor class has a distinct spectral tuning that overlaps with the others. Using a multiprimary system, in which the primaries (i.e. LEDs) have different spectra, it is possible to target one class of photoreceptors while maintaining the others at a constant activity level, effectively silencing them [@Shapiro1996; @Spitschan2018, this paper also offers a clear explanation of how to implement silent substitution]. We calculated silent substitution solutions using the _PySilSub_ toolbox [@Martin2023], using linear algebra. The outputs of the two light engines (see Figure \@ref(fig:spectraplots)b,c) were calibrated using an Ocean Optics Jaz spectroradiometer, which was wavelength-calibrated to an Argon lamp and intensity calibrated using a NIST-traceable light source. Each primary was also linearized using a polynomial fit (see Figure \@ref(fig:primarylinearity) for details). We used the 10-degree cone fundamentals [@Stockman2000], and estimates of melanopsin absorbance spectra from CIE S 026 [discussed in @Martin2023] to calculate $\alpha$-opic irradiance.

@@

The diffuser discs were positioned in the objective planes of the lenses of a modified VR headset (SHINECON SC-G01, Dongguan Shinecon Industrial Co. Ltd., Guangdong, China), which was used by the participants to view the stimuli. The stimuli were two discs of flickering light with a diameter of approximately 30 degrees, which were fused together into a cyclopean percept resembling a donut-shaped ring of light, similar to that used in other studies [e.g., @Barrionuevo2016; @Murray2018; @Spitschan2014; @Spitschan2015; @Zele2018]. The VR headset modifications allowed for small adjustments to account for individual differences in interpupillary distance and focal length. The use of this set up allowed us to modulate the stimuli in three different ocular configurations, similar to the ones used in a previous study [@Segala2023]: monocular, binocular and dichoptic. In the monocular configuration, the unstimulated eye still saw a non-flickering disc of mean light flux. A schematic of the stimulation system is shown in Figure \@ref(fig:spectraplots)a. Pupillometry data were collected using a binocular Pupil Core eye-tracker headset (Pupil Labs GmbH, Berlin, Germany @Kassner2014) running at 120 Hz, and the signals were recorded with the Pupil Capture software.

A previous study [@Segala2023] used a temporal frequency of 2Hz for foveal luminance flicker, and recorded EEG data simultaneously with pupillometry. Initial pilot experiments indicated that this frequency was too high to elicit measurable responses when stimulating individual photoreceptor pathways. For all experiments, we therefore used a primary flicker frequency of 0.5Hz, as previous literature showed that this was slow enough to elicit a pupil response from all photoreceptor classes [@Spitschan2014]. We also focussed on only recording pupillometry data as this frequency would be too slow to elicit steady-state EEG responses [@Norcia2015].

@@

Before the start of each experiment, participants adjusted the objective planes of the lenses with the help of the experimenter until the stimulus was in focus and they perceived the two pieces of blackout material as one fused disc. Pupil responses to binocular temporal contrast modulations were examined in a factorial design that combined six ocular conditions and five temporal contrast levels: 6, 12, 24, 48 and 96% of the available dynamic range. This design, similar to that used in previous studies [@Segala2023; @Baker2017; @Baker2020], was applied in four separate experiments, each with a different mode of photoreceptor stimulation. In the first three conditions, the discs flickered at 0.5 Hz, in either a monocular, binocular or dichoptic arrangement. In the dichoptic condition the non-target eye saw a flickering fixed contrast of 48% of the available dynamic range. In the remaining three conditions (the cross-frequency conditions) one eye’s disc flickered at 0.4 Hz, and the other eye’s disc flickered at 0.5 Hz. This included monocular responses at 0.4 Hz, as well as binocular (one eye sees each frequency at the target contrast) and dichoptic (target stimulus flickering at 0.5 Hz, mask contrast of 48% at 0.4 Hz in the other eye) arrangements. We counterbalanced presentation of the target stimulus across the left and right eyes.

@@

The pupillometry data were analysed using the same method used in a previous study [@Segala2023]. The data were converted from mp4 videos to a csv text file using the Pupil Player software [@Kassner2014], which estimated pupil diameter for each eye on each frame using a 3D model of the eyeball. The individual data were then loaded into R for analysis, where a ten-second waveform for each trial in each eye was extracted (excluding the first two seconds after stimulus onset). We interpolated across any dropped or missing frames to ensure regular and continuous sampling over time. The Fourier transform was calculated for each waveform, and all repetitions of each condition were pooled across eye and then averaged. Finally, data were averaged across all participants to obtain the group results. We used coherent averaging and at each stage we excluded data points with a Mahalanobis distance exceeding $D$ = `r SDthresh` from the complex-valued mean [@Baker2021]. For monocular stimulation, we confirmed that the consensual response was equivalent to the response in the stimulated eye.

@@

To quantitatively summarise our data, we used the same model described in a previous study [@Segala2023]. The model has the same general form as the first stage of the contrast gain control model proposed by Meese and colleagues [@Meese2006], but omits the second stage. For the previous model that was used by Segala and colleagues [@Segala2023], the exponent of the numerator and denominator had fixed values of 2 and (implicitly) 1. Here, we allow these parameters (called _p_ and _q_) to be free, in order to permit different shapes of contrast response function, e.g. accelerating or saturating. The responses of the left eye and right eye channels are as follows:

@@

A recent psychophysical work [@Baker2024] has looked at binocular interactions in the L-M and S-cone pathways, and compared these to the light flux pathway. Using a contrast discrimination paradigm with spatial modulations of light flux and colour, the study [@Baker2024] found equally strong interocular suppression in all three pathways. This is rather different from our pupillometry results here, which demonstrate weaker suppression in the light flux and S-cone pathways than in the L-M pathway (see Figure \@ref(fig:FirstHarmonicPlots)). However, the current experiments involve temporal modulations, which are quite distinct from the spatial modulations used in the study by Baker and colleagues [@Baker2024]. Other recent work [@Segala2023] has shown that temporal luminance modulations involve much weaker interocular suppression in the cortical response than do spatial luminance modulations [@Baker2017]. These different normalization processes might reflect different priorities for spatial and temporal vision. Spatial vision aims to fuse images to provide binocular single vision, and benefits from 'ocularity invariance' [@Baker2007], in which visual appearance is constant when viewed with one eye or two. Temporal vision is critical for motion perception, which can involve alternative binocular computations, such as calculating velocity differences between the eyes [@Kaestner2019]. Of course the present measurements were of pupil size, which may be subject to different anatomical and functional constraints from those in the cortex. Future research should aim to extend our current findings to perception using psychophysical approaches.

@@
